\proglang{EpiCompare}: Working title

### 2020-04-22 jss/jss-article-rnw/article.Rnw
Added R0 section. yay

@@

%% - \Abstract{} almost as usual
\Abstract{
  This short article illustrates how to write a manuscript for the
  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
  it is recommended to keep the {\LaTeX} code as simple as possible,
  i.e., avoid inclusion of packages/commands that are not necessary.
  For outlining the typical structure of a JSS article some brief text snippets
  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
  discussing count data regression in \proglang{R}. Editorial comments and
  instructions are marked by vertical bars.
}

@@

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Achim Zeileis\\
  Journal of Statistical Software\\
  \emph{and}\\
  Department of Statistics\\
  Faculty of Economics and Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstr.~15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
}

@@

>>>>

\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}

@@

%% - \Abstract{} almost as usual
\Abstract{
We do something
}

@@

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Shanon K. Gallagher\\
  Biostatistics Research Branch\\
  Division of Clinical Research\\
  National Institute of Allergy and Infectious Diseases\\
  5601 Fishers Lane\\
  Rockville, MD 20852\\
  E-mail: \email{shannon.gallagher@nih.gov}
}

@@

\SweaveInput{chapters/app1-proofs.Rnw}

### 2020-04-22 jss/jss-article-rnw/chapters/ch1-intro.Rnw
Added R0 section. yay

The recent (and currently on-going) COVID-19 global epidemic has galvenized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area \cite{FILL IN}. Infectious disease models models attempt, typically, to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{biggerstaff2016}; or 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \cite{roberts2004}. The COVID-19 epidemic highlights how both goals are very important, and that preventing and mitigating the future spread of infectious disease which inflicts enormous burdens on the world: millions of lives lost and billions of dollars spent yearly, is so important.

The current epidemic has also highlighted that these models are just one piece of the overall analysis pipeline. John Hopkin's \_\_\_ Dashboard and county, state and national tabulation of rudimentary numerical summaries emphasis the need to strong data to assess the spread of infectious disease (multiple \cite{FILL IN}?). Stilly, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individal / agent level, network structure or just aggregate flows (review paper \cite{FILL IN}?). All placing individuals / proportions of the populations into different states (e.g. suspectible, exposed, infected, recovered, etc.). With all these models, review and comparison papers in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center helps the individual practioner decide the correct approach.

@@

More specifically, \pkg{EpiCompare} provides epidemiologists a toolkit containing:
\begin{itemize}
  \item \textbf{Visualization tools} to visualize three-state epidemics and simulations from three-state\footnote{The most common three-state model that will be examined is the SIR model (with the three states: Susceptible, Infected and Recovered), and will be discussed in more detail in Section \ref{}.} models in a time-free manner using ternary plots and prediction bands. For agent-based three-state models we also provide visualization tools to let the user easily explore how different characteristics of the agents relate to different experiences in the epidemic.
  \item General \textbf{comparison tools} to compare epidemics and epidemic models that have higher numbers of states (again in a time-free manner), allowing for the user to examine the differences between models through simulations, and if an epidemic is similar to a model through simulations and prediction bands.
  \item \textbf{Conversion tools} to:
    \begin{itemize}
      \item Convert and then compare models from \textit{standard \proglang{R} epidemic packages} as well as internal agent-based models, and epidemics in a common framework.
      \item Convert \textit{agent-based information into aggregate} to compare in the aggregate framework described above.
Installation
    \end{itemize}
\end{itemize}

>>>>

The recent (and currently on-going) COVID-19 global epidemic has galvanized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious disease models models attempt, typically, to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{biggerstaff2016}; or 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \cite{roberts2004}.  A third use of these models is to 3) forecast or project scenarios under pre-specified parameters (i.e. 'what if's) \citep{}. The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.  Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.  Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks.

The current epidemic of COVID-19 also highlights that these models are just one piece of the overall analysis pipeline. John Hopkin's \_\_\_ Dashboard and county, state and national tabulation of reported cases and deaths emphasize the need to strong data to assess the spread of infectious disease (multiple \cite{FILL IN}?). Stilly, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individal / agent level, network structure or just aggregate flows (review paper \cite{FILL IN}?). All placing individuals / proportions of the populations into different states (e.g. suspectible, exposed, infected, recovered, etc.). With all these models, review and comparison papers in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center helps the individual practioner decide the correct approach.

@@

The goal of \pkg{EpicCompare} is to suplement and enhance current infectious disease modeling pipelines as well as to encourage comparisons across these pipelines.
More specifically, \pkg{EpiCompare} provides epidemiologists a toolkit containing:
\begin{itemize}
  \item \textbf{Visualization tools} to visualize three-state epidemics and simulations from three-state\footnote{The most common three-state model that will be examined is the SIR model (with the three states: Susceptible, Infected and Recovered), and will be discussed in more detail in Section \ref{}.} models in a time-free manner using ternary plots and prediction bands. For agent-based three-state models we also provide visualization tools to let the user easily explore how different characteristics of the agents relate to different experiences in the epidemic.
  \item General \textbf{comparison tools} to compare epidemics and epidemic models that have higher numbers of states (again in a time-free manner), allowing for the user to examine the differences between models through simulations, and if an epidemic is similar to a model through simulations and prediction bands.
  \item \textbf{Conversion tools} to:
    \begin{itemize}
      \item Convert and then compare models from \textit{standard \proglang{R} epidemic packages} as well as internal agent-based models, and epidemics in a common framework.
      \item Convert \textit{agent-based information into aggregate} to compare in the aggregate framework described above.
Installation
    \end{itemize}
\end{itemize}

### 2020-04-22 jss/jss-article-rnw/chapters/ch2-vis.Rnw
Added R0 section. yay

>>>>

One issue that the COVID-19 outbreak has brought to the forefront is the importance of figures and graphs to inform, support, and even persuade lay people of technical and scientific results \cite{}.  Quantities and trends shown in these visualizations include exponential growth curves, S-curves, peak day and percent of infectious indviduals, ICU capacity, and deaths over time and uncertainty of such estimates.

\pkg{EpiCompare} supports traditional visualizations (e.g. \# in condition vs. time) and also introduces a novel visualization of epidemics via time free analysis.  In particular, we use ternary plots that show Susceptible-Infectious-Recovered data/models (SIR).  The advantages of these ternary plots are 1) a standard way to compare epidemics with different outbreak times and time scales, 2) a novel interpretation of $R_0$, the initial reproduction number under some circumstances, and  3) clear visualization of joint SIR uncertainty via conformal bands.

Ternary plots are sometimes used in chemistry \citep{} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).  Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.  The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{}), and also happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.

Ternary plots are time-free in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g. days vs. years) or of SIR data of the same disease in different areas (our first case study in Sec. \ref{sec:ex1} focuses on this).

Ternary plots also, under certain conditions, have a special relationship with $R_0$, the initial reproduction number, which has been called the ``most important quantity in epidemiology'' \citep{}.  The reproduction is itself a time-free concept and is the average number of secondary infections primary infection will generate when the primary infection is added to a completely susceptible population.  To see the importance of $R_0$, one need only read the newspaper \citep{} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{}.

In regards to traditional $X$ vs. time plots, $R_0$ is difficult to visualize.  For example, as seen in Fig. \ref{fig:r0-trad}, it is possible that two models have the same value of $R_0$ but there is no way of knowing that from looking at the graph.

It is possible to identify, at a glance, which epidemic paths have the same value of $R_0$ via ternary plot visualizations, under some circumstances.  To show, this we first need to introduce the classic, deterministic Kermack and McKendrick SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.  The movement of individuals among the states is given by the following differential equations (Eq. \eqref{eq:sir-ode}) where $\beta$ is the average infection rate and $\gamma$ is the average recovery rate,
\begin{align}\label{eq:sir-ode}
      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
      R^\prime(t) &= \gamma I(t) \nonumber.
  \end{align}

If we have two SIR models that follow the Kermack and McKendrick SIR model, have the same percentage of individuals in the initial states, and the same value of $R_0$ then the two models will create exactly overlapping paths in a ternary plot.

More formally, let two Kermack and McKendrick (see Eq. \eqref{eq:sir-ode}) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$. Assume both models have initial values $(S(0), I(0), R(0))$.  Let $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.  Equivalently, $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$ for some $a > 0$.

\begin{theorem}\label{thm:sir-scale}
Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{} and is shown in detail in Appendix \ref{app:proof}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.

### 2020-04-22 jss/jss-article-rnw/chapters/ch5-ex1.Rnw
Added R0 section. yay

>>>>

%%% Move to example?  Think I will
As a motivating example for time free analysis visualizations, we would like to compare the outbreak of COVID-19 across different states.  Typically, the raw data must be adjusted (or stochastically imputed) in order for \textit{fair} comparisons to be seen.  A fairly common example of this is when visualizatios show deaths or infections by percent of the area's population.  Another issue is that outbreaks differ temporally:  the first state (CA?) with confirmed COVID cases occurred on January X while the final state (WV?) occurred on March Y.  One approach to adjust for the different outbreak times is to show the number of cumulative cases versus the time since the first reported outbreak in the state.  However, this can result in unexpected features such as shown in Figure \ref{fig:covid-states-time-adjusted}, where Massachusetts has an early reported case but does not have a sustained outbreak until long after that first case.  Instead of adjusting temporally to the first case, we could instead adjust to when an threshold of cases (raw or percentage) is determined but how this threshold is determined is unclear.  Time free analysis
